\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2021-08-05 move_schedules.tex
various edits throughout the main body of the paper

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Ordered set of filtrations $\mathcal{K}$, each of size $m$, with bijections $f_{i}: K_i \to K_{i+1}$
    	\Ensure Valid $R = D V$ decompositions are computed for each filtration $K_1, K_2, \dots, K_n$
    	\Procedure{SparseSchedule}{$\mathcal{K} = ( K_1, K_2, \dots, K_n )$} 
    		\State $\mathcal{S} = \emptyset$
    		\For{$i = 1$\textbf{ to }$n-1$}
    		    \State $(p, q) \gets ([m], \, \mathrm{Im}\, f^\ast_{i})$ 
    			\State $\mathrm{lis}_q \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
    			\State $\mathcal{S} \gets \mathcal{S} \cup \mathrm{GreedySchedule}(p, q, \mathrm{lis}_q)$ \Comment{$O(d\log d)$, see section~\ref{sec:proxy_objective}}
    		\EndFor
	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial K_1$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$}
	    		\If{$i < j$}
	    			\State $(R, V) \gets$ \Call{MoveRight}{i, j}
	    		\Else
	    			\State $(R, V) \gets$ \Call{MoveLeft}{i, j}
	    		\EndIf
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, 
and is meant to serve as a guidepost before discussing its other components in depth. 
Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. 
Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly. Due to the combinatorial aspect of our approach, there is actually no need to have access to---or explicitly store---the entire family of filtrations and the schedules between them. The entire approach can be performed online, keeping at most two filtrations and one decomposition in memory at any given time.

@@

\subsubsection{The Discrete Setting} 
Since moves do not maintain a valid decomposition during movement, it is worth asking if we can achieve better bounds than vineyards, by removing the assumption that the goal is to simulate persistence across a supplied homotopy $F(x, \tau) = \mathbb{X} \times [0,1] \to \mathbb{R}$. In the case of vineyards, the bound is clear: whether assuming a homotopy between a pair of filtrations $(K_0, K_1)$ or not, if $\kappa = K_\tau(K_0, K_1)$, then $\Omega(\kappa)$ adjacent transpositions are required to transform $K_0 \mapsto K_1$. 
Allowing move operations in any order, we have $O(m)$ as a trivial upper bound: simply move each simplex in $K_0$ into its position in the filtration given by $K_1$ in the order given by the latter. However, it's not immediately clear whether this bound is tight. To establish this formally, we require a few definitions.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. It was shown by Baik et al.~\cite{} that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c \approx -1.77108$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$, it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. Since this lower bound grows similarly to $m!$ when $t \approx O(m)$ and $d << m$ fixed, we expect $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
For example, the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $m=417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.

>>>>

\begin{algorithm}[h]
	\caption{Scheduling algorithm}\label{alg:schedule}
	\setstretch{1.05}
    \begin{algorithmic}[1]
    	\Require Ordered set of filtrations $\mathcal{K}$, each of size $m$, with bijections $f_{i}: K_i \to K_{i+1}$
    	\Ensure Valid $R = D V$ decompositions are computed for each filtration $K_1, K_2, \dots, K_n$
    	\Procedure{SparseSchedule}{$\mathcal{K} = ( K_1, K_2, \dots, K_n )$} 
    		\State $\mathcal{S} = \emptyset$
    		\For{$i = 1$\textbf{ to }$n-1$}
    		    \State $(p, q) \gets ([m], \, \mathrm{Im}\, f^\ast_{i})$ 
    			\State $\mathrm{lis}_q \gets \mathrm{LIS}(q)$ \Comment{$O(m \log \log m)$}
    			\State $\mathcal{S} \gets \mathcal{S} \cup \mathrm{GreedySchedule}(p, q, \mathrm{lis}_q)$ \Comment{$O(d\log d)$, see section~\ref{sec:proxy_objective}}
    		\EndFor
	    	\State $(R, V) \gets$ \Call{Reduction}{$D = \partial K_1$}
	    	\For{$(i,j)$\textbf{ in }$\mathcal{S}$}
	    		\If{$i < j$}
	    			\State $(R, V) \gets$ \Call{MoveRight}{i, j}
	    		\Else
	    			\State $(R, V) \gets$ \Call{MoveLeft}{i, j}
	    		\EndIf
	    	\EndFor
    	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\noindent Algorithm~\ref{alg:schedule} is purely illustrative at this point, 
and is meant to serve as a guidepost before discussing its other components in depth. 
Specifically, lines (4-5) are discussed next in section~\ref{sec:schedule_sizes}, and the algorithm for line (6) is given in section~\ref{sec:schedule_cost}. 
Note that the loop starting at line (3) in Algorithm~\ref{alg:schedule} could always be merged with the loop starting at line (8), i.e. the individual schedules $S_i$ may be computed on the fly. Due to the combinatorial aspect of our approach, there is actually no need to have access to---or explicitly store---the entire family of filtrations and the schedules between them. That is, Algorithm~\ref{alg:schedule} can be easily modified to be completely online, keeping at most two filtrations and one decomposition in memory at any given time.

@@

\subsubsection{The Discrete Setting} 
Since moves do not maintain a valid decomposition during movement, it is worth asking if we can achieve better bounds than vineyards, by removing the assumption that the goal is to simulate persistence across a supplied homotopy $F(x, \tau) = \mathbb{X} \times [0,1] \to \mathbb{R}$. In the case of vineyards, the bound is clear: whether assuming a homotopy between a pair of filtrations $(K_0, K_1)$ or not, if $\kappa = K_\tau(K_0, K_1)$, then $\Omega(\kappa)$ adjacent transpositions are required to transform $K_0 \mapsto K_1$. 
In contrast, if we allow move operations in any order, we have $O(m)$ as a trivial upper bound: simply move each simplex in $K_0$ into its position in the filtration given by $K_1$ in the order given by the latter. However, it's not immediately clear whether this bound is tight. To establish this formally, we require a few definitions.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. The seminal work by Baik et al.~\cite{baik1999distribution} established that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c = -1.77108...$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$, it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. Since this lower bound grows similarly to $m!$ when $t \sim O(m)$ and $d << m$ fixed, we expect $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
For example, the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $m=417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.
